Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus

### 2024-11-14 paper/paper.tex
progress on sum eta
related work

Prior approaches to verified parsing focus on verification of a
particular grammar formalism such as non-left-recursive grammars or LL(1) grammars \cite{lasserCoStarVerifiedALL2021,EdelmannZippy2020,danielssonTotalParserCombinators2010}.
\steven{These have restrictions on left recursion or LL(1)-ness. Is it the most accurate
  to call that a difference in formalism? It would seem that these are all CFGs
  + conditions. Maybe these aren't the best works to
  compare to if they're verifying particular algorithms}
Each new grammar formalism must then be extended with its own verified
implementation.

@@

\subsubsection{Lambek Calculus and Categorial Grammar}
Lambek's original syntactic calculus \cite{lambek58} describes a logical system
for linguistic derivations. Later, he explicitly gives semantics to his
syntactic calculus inside of a non-commutative biclosed monoidal category
\cite{lambek1988categorial}. Following Lambek's illuminating ideas, Buszkowski
gives a succinct overview of the
rich body of work studying the role of linear logic and lambda calculi in
linguistics \cite{buszkowskiTypeLogicsGrammar2003}. \steven{Is citing a survey
  like this okay?} Grammars based on logic in
this manner may go by the names Lambek calculus, typelogical grammar, or
categorial grammar. Such categorial grammars correspond to the subtheory of our
calculus generated by $\otimes$, $\lto$, and $\tol$.

@@

Luo \cite{luo} describes a Lambek calculus with dependent types. However, they
do not provide an analagous denotational semantics, so it is not clear what
exactly their semantics are. Further, their contexts of non-linear variables are
ordered, while ours are not, so they provide a distinct dependent extension
whose types do not have a one-to-one correspondence with the dependent types in \theoryabbv.

@@

Krishnaswami and Yallop \cite{krishnaswami_typed_2019} provide a type system for
demonstrating that carries proves of the LL(1) condition for context-free
grammars. Despite being a type system over formal grammars, their interpretation
of types is very different from the semantics of \theoryabbc.

@@

\paragraph{$\bigoplus\eta$}
\steven{TODO $\eta$ for sums}

>>>>

Prior approaches to verified parsing focus on verification of a
particular grammar formalism such as non-left-recursive grammars or
LL(1) grammars
\cite{lasserCoStarVerifiedALL2021,EdelmannZippy2020,danielssonTotalParserCombinators2010}.
Each new grammar formalism is extended with its own independent
verified implementation.

@@

\subsubsection{Lambek Calculus and Categorial Grammar}
Lambek's original syntactic calculus \cite{lambek58} describes a logical system
for linguistic derivations. Later, he explicitly gives semantics to his
syntactic calculus inside of a non-commutative biclosed monoidal category
\cite{lambek1988categorial}. This led to many uses of non-commutative linear logic and lambda calculi in
linguistics \cite{buszkowskiTypeLogicsGrammar2003}. Grammars based on logic in
this manner may go by the names Lambek calculus, typelogical grammar, or
categorial grammar. Such categorial grammars correspond to the subtheory of our
calculus generated by $\otimes$, $\lto$, and $\tol$.

@@

The most similar prior work to our own is Luo's Lambek calculus with
dependent types \cite{luo}. Their design differs from our own in
allowing linear types to depend on other linear types and supporting
directed $\Pi$ and $\Sigma$ types that have no analog in our
system. They do not provide a semantics for these connectives, and it
is unclear how to interpret their connectives in our grammar
semantics. Further, we provide several examples showing that
\theoryabbv is a practical system for describing grammar formalisms
and parsers, and it is unclear if these could be implemented in their calculus.

@@

Krishnaswami and Yallop \cite{krishnaswami_typed_2019} provide a type system for
demonstrating that carries proves of the LL(1) condition for context-free
grammars. Despite being a type system over formal grammars, their interpretation
of types is very different from the semantics of \theoryabbv.

@@

\paragraph{$\bigoplus\eta$}
It suffices to show
\[
  \semg{\letin {\sigma~x~a} {c'} {f[\sigma~x~a/c]}} = \semg{f[c'/c]}
\]
\begin{proof}
  \begin{align*}
    \semg{\letin {\sigma~x~a} {c'} {f[\sigma~x~a/c]}}
    &= [ \sem{f[\sigma~x~a/c]} ]_{(x : X)} \circ \coprod_{x:X}(m^{-1}) \circ d \circ (id \otimes \lambda) \otimes id \circ m \otimes id \circ m
  \end{align*}
\end{proof}
\max{todo: finish}
